---
title: "Rheumatoid arthritis meta-analysis - Batch effect correction"
author: "Madhulika Mishra, Lucas Barck, Pablo Moreno, Guillaume Heger, Yuyao Song, Janet M. Thornton, Irene Papatheodorou"

output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 4
editor_options: 
  markdown: 
    wrap: 72
---

## Introduction
In this workflow, I want to perform the batch effect correction on datasets that Madhulika Mishra and I selected for our meta-analysis on osteoarthritis:

* **E-GEOD-1919**: 5x control; 5x OA
* **E-GEOD-55235**: 10x control; 10x OA
* **E-MTAB-5564**: 10x control; 14x OA
* **E-GEOD-51588**: 10x control; 40x OA

## Batch effect correction
#### Import of the packages and datasets

```{r}
# Loading the packages
library(magrittr)
library(purrr)
library(dplyr)
library(SelectBCM)
library(Biobase)
library(matrixStats)
library(S4Vectors)
library(ComplexHeatmap)
library(tictoc)
library(tidyverse)

# Loading the experiments as ExpressionSet or SummarizedExperiment (folder is only allowed to contain .RData files)
experiments <- load_experiments('/Users/lucasbarck/Desktop/EMBL_EBI/Lucas_Barck_summer_internship_2021/4_OA_meta_analysis/output/Preprocessing/unnormalized data')
```

#### Check-up of pData
```{r}
unique(pData(experiments$E_GEOD_51588_eset_bgc.RData)$disease)
unique(pData(experiments$E_GEOD_1919_eset_bgc.RData)$disease)
unique(pData(experiments$E_GEOD_55235_eset_bgc.RData)$disease)
unique(pData(experiments$E_MTAB_5564_eset_bgc.RData)$disease)
```

#### Removement of isolated experiments
```{r}
experiments %<>% remove_isolated_experiments('disease')

sapply(experiments,dim)

sep_experiments <- experiments
```
Note that the feature numbers range from 8604 to 20744.

#### UpSet plot
The UpSet plot is a very intuitive way to estimate the gene overlap between different chips. Here, I will use it to analyse the compatibility of the different chips.
```{r}
listInput <- sapply(sep_experiments, featureNames)
combination_matrix <- make_comb_mat(listInput, mode = "intersect")
combination_matrix <- combination_matrix[comb_degree(combination_matrix) >= 3]
ComplexHeatmap::UpSet(combination_matrix, 
                      comb_col = "black",
                      pt_size = unit(2.5, "mm"), lwd = 2,
                      bg_col = "#F0F0F0", bg_pt_col = "#CCCCCC",
                      row_names_side = "left",
                      row_names_gp = grid::gpar(fontsize = 8),
                      comb_order = order(comb_size(combination_matrix), decreasing = TRUE),
                      
                      top_annotation = HeatmapAnnotation(
                      degree = as.character(comb_degree(combination_matrix)),
                      "Intersection\nsize" = anno_barplot(comb_size(combination_matrix), 
                      border = FALSE, 
                      gp = gpar(fill = "darkgrey"), 
                      height = unit(2, "cm")), 
                      annotation_name_side = "left", 
                      annotation_name_rot = 0),
                      
                      
                      right_annotation = upset_right_annotation(combination_matrix, 
                      ylim = c(0, 22000),
                      gp = gpar(fill = "darkgrey"),
                      annotation_name_side = "bottom",
                      axis_param = list(side = "bottom"))
                      )

sapply(sep_experiments,dim)
```
The UpSet plot shows that the maximum number of gene overlap is achieved by excluding E-GEOD-1919. The reason for that is the comparably small number of total genes in the first place (8604). \
However, the gene overlap between all four datasets is arguably very good. Therefore, I will keep every dataset.


#### Merging experiments in a single dataset
```{r}
experiments %<>% merge_experiment.ExpressionSet(log = FALSE)
dim(experiments)
```
The merge_experiments function reduced the number of features to 7926. <br/><br/>

There are still rheumatoid arthritis samples in the data frame. These samples need to be excluded since this meta analysis focuses on the difference between normal and osteoarthritis.

```{r}
# Isolating control and OA samples
unique(pData(experiments)$disease)
filter <- colnames(experiments)[pData(experiments)$disease != "rheumatoid arthritis"]
experiments <- experiments[, filter]
table(pData(experiments)$disease)
dim(experiments)

pData(experiments) %>% select(disease, batch) %>% group_by(batch) %>% table()
```

```{r}
# Export of the merged file
save(experiments, file="osteoarthritis_merged_datasets.RData")
```


#### Batch effect detection
```{r}
batch_effect.raw <- detect_effect(experiments,experiment = experiments, batch.factors=c("batch","disease"),10,10,'entrezgene_id')
Result <- do.call(rbind, lapply(batch_effect.raw, data.frame))
Result
```
These results indicate that there is a strong batch effect present in the data.


#### Batch effect correction
```{r}
fData(experiments)$entrezgene_id <- rownames(fData(experiments))

tic("batch_correction")
result <- batch_correction(experiment= experiments,model=~disease, batch = "batch",filter='entrezgene_id')
toc()
summary(result)

OA_batch_res_with_NAs <- result

```


#### Assessment of batch-correction methods
Some batch effect correction methods created NAs in the data. This issue causes `batch_evaluation()` to fail, and is also problematic for later analysis. Therefore, it is necessary to identify the respective datasets and substitute the NAs with "0".
```{r}
# Identify affected datasets
for(i in 1:length(result)){
  print(names(result)[i])
  print(all(is.finite(exprs(result[[i]]))))
}
```
These methods that produced NAs are: data.quantile, Q_ComBat, data.Q_naiveRandRUV_HK, and data.Q_naiveRandRUV_empi.controls.

```{r}
# Substitute NAs
exprs(result$data.Q_naiveRandRUV_HK)[is.na(exprs(result$data.Q_naiveRandRUV_HK))] <- 0
exprs(result$data.Q_naiveRandRUV_empi.controls)[is.na(exprs(result$data.Q_naiveRandRUV_empi.controls))] <- 0
exprs(result$data.quantile)[is.na(exprs(result$data.quantile))] <- 0
exprs(result$data.Q_ComBat)[is.na(exprs(result$data.Q_ComBat))] <- 0

osteoarthritis_batch_result <- result
save(osteoarthritis_batch_result, file="osteoarthritis_batch_cor_result.RData")
```


Since the gender information is not available for every experiment, the column should be removed before the batch effect correction evaluation.
```{r}
# Remove gender column
for (i in 1:length(result)){
  pData(result[[i]]) <- pData(result[[i]]) %>% select(-sex) %>% select(-FactorValue..SEX.)
}

pData(experiments) <- pData(experiments) %>% select(-sex) %>% select(-FactorValue..SEX.)
```


```{r}
# Batch effect correction evaluation
assessment <- SelectBCM::batch_evaluation.microarray(result, batch.factors=c("batch","disease"), experiments,10,10,filter='entrezgene_id')
assessment

final <- diagnostic_matrix.microarray(assessment)

dia_plot <-  SelectBCM::diagnostic_plot.microarray(final)

svg("Diagnostic_plot_OA_example.svg")

ggarrange(dia_plot$diagnostic_plot,dia_plot$diagnostic_stat_summary,
          labels = c("a.", "b."),
          legend = "right", ncol = 1, nrow = 2
)

dev.off()


bcm_ranking(final)
```

Therefore, the top performing batch effect correction methods are:

1) ComBat2
2) Q_ComBat
3) ComBat1


# Exporting R session and session info
```{r}
# Saving the R session
save.image(file = "osteoarthritis_batch_cor_session.RData")
```


<details>
<summary>
**See for session info**
</summary>
```{r}
sessionInfo()
```
</details>
